High geomagnetic field intensity in the late Mesoproterozoic recorded by Midcontinent Rift anorthosite xenoliths

### 2021-08-28 manuscript/Manuscript.tex
manuscript text udpate

\begin{keypoints}
\item New ca. 1092 Ma Midcontinent Rift anorthosite record high paleointensity values
\item High VADM of the geomagnetic field during the late Mesoproterozoic is inconsistent with previous interpretation of long term decay
\item A strong geodynamo like today likely persisted throughout 25 myr during the Midcontinent Rift development
\end{keypoints}

% keywords: absolute paleointensity, Laurentia, Midcontinent Rift, anorthosite,

\begin{abstract}
High-quality absolute paleointensity

\end{abstract}
Retrieving paleomagnetic information from Precambrian rocks can aid in our understanding of the long-term evolution of the geodynamo. A recent paleointensity study by \cite{Sprain2018a} reported a ca. 1.1 Ga paleomagnetic field strength similar to today based on the volcanics of the Midcontinent Rift System (MRS). That study suggested that this result is consistent with models wherein this time period of the Proterozoic is characterized by a strong rather than weak geomagnetic field. However, to further evaluating the evolution of the geomagnetic field during the Precambrian, more paleointensity data are needed.

@@

@@

@@

\section*{Introduction}

@@

\bibliography{}

>>>>

\begin{keypoints}
\item New ca. 1092 Ma Midcontinent Rift anorthosite record high paleointensity values
\item High VADM of the geomagnetic field during the late Mesoproterozoic is inconsistent with previous interpretation of long term decay
\item A strong geodynamo like today likely persisted for about 25 myr during the Midcontinent Rift development
\end{keypoints}

% keywords: absolute paleointensity, Laurentia, Midcontinent Rift, anorthosite, geodynamo

\begin{abstract}
New high-quality absolute paleointensity data from anorthosite xenoliths within the Beaver River diabase intrusions of the North American Midcontinent Rift (MCR) record a virtual axial dipole moment (VADM) estimate of $\sim$80 ZAm$^2$ ca. 1092 Ma. This cooling rate corrected paleointensity estimate from an MCR intrusive unit is similar to previous results from extrusive units that span from ca 1108 Ma to ca. 1085 Ma. Taken together, new and previously published paleointensity data from the Midcontinent Rift support that a geomagnetic field strength persisted through the Midcontinent Rift development and the data are inconsistent with the interpretation of a decay of Earth's geodynamo strength through the late Mesoproterozoic.

\end{abstract}

\section*{Introduction}
%The question of the core and the importance of the observational data as complementary for numerical simulation work.

That we have a solid inner core today is the result of secular cooling and differentiation of the layered Earth. The timing of inner core formation during Earth's thermal evolution is amongst the major unsolved questions in Earth's history. Due to the uncertain constraints on the thermal conductivity of the core \cite{Konopkova2016a, Gomi2013a, Ohta2016a}, numerical simulations have given drastically different predictions on the inner core nucleation age, ranging from about half a billion years ago to prior to the Archean \cite{Pozzo2012a, Gubbins2004a, Nimmo2015a}. Modeling has suggested that the emergence of the solid inner core could have expressed as changes of Earth's surface magnetic field strength due to the onset of compositional convection between the solid inner core and the liquid outer core \cite<e.g.>{Aubert2009a}. Thus, absolute paleointensity records preserved by magnetic minerals in igneous rocks may provide us with a probe into the birth of the inner core. Acquiring these observational data is crucial for advancing our understanding of Earth's thermal evolution and for testing existing models on the commencement of the inner core growth. However, due to the scarcity of well-preserved ancient rocks and low success rates typical of paleointensity experiments, only 9\% of the records in the current global paleointensity database (\href{http://earth.liv.ac.uk/pint/}{PINT, \cite{Biggin2010a}}) are from before the Cambrian and the existing records sparsely span over 3 billion years of time (Fig. \ref{fig:all_PINT_data}).

\citeA{Bono2019a} filtered the available Precambrian paleointensity database (PINT) based on a set of quality criteria. A polynomial fit through the resulting paleointensity compilation from the Archean and the Proterozoic was used to interpret that a decrease trend of the estimated geomagnetic dipole strength occurred before the inner core started to form. In addition, \citeA{Bono2019a} recovered very low paleointensity values from ca. 565 Ma single silicate crystals  that were interpreted to be consistent with it reflecting the weak geodynamo after progressive cooling of the core before the onset of the rigorous compositional convection driven by the nucleation of the solid inner core. Such decay trend is recently supported by new low paleointensity estimates derived from the ca. 720 Ma Franklin Large Igneous Province \cite{Lloyd2021a}.

However, previous paleointensity data from the MCR rock units \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record VADM values distinctly higher than what is predicted by the long term decay curve. These rock units notably are extrusive volcanics and span a tme period of about 25 myr during the development of the Midcontinent Rift magmatism. Many previous results from the Midcontinent Rift show non-ideal double-slope paleointensity behavior, which are difficult to interpret as they can yield two paleointensity estimates: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. Nevertheless, single-slope, ideal behavior from the Lake Shore Traps \cite{Kulakov2013a} lava flows passed the selection criteria of \citeA{Bono2019a}, and a few lava flows yielded single-slope behavior from the Osler Volcanics \cite{Sprain2018a}. These data are inconsistent with the interpretation of a low geomagnetic field strength during the Midcontinent Rift because they yield significantly higher VADM estimates than that projected by the trend from \citeA{Bono2019a}.

To further enrich the Precambrian paleointensity database, and to test the hypothesis that the Earth's geomagnetic dipole strength remained at high values throughout the Midcontinent Rift, in this study we perform ``IZZI" paleointensity experiments \cite{Tauxe2004a} on a suite of anorthosite xenoliths hosted by the intrusive Beaver River diabase of the Beaver Bay Complex within the Midcontinent Rift.

%Retrieving paleomagnetic information from Precambrian rocks can aid in our understanding of the long-term evolution of the geodynamo. A recent paleointensity study by \cite{Sprain2018a} reported a ca. 1.1 Ga paleomagnetic field strength similar to today based on the volcanics of the Midcontinent Rift System (MRS). That study suggested that this result is consistent with models wherein this time period of the Proterozoic is characterized by a strong rather than weak geomagnetic field. However, to further evaluating the evolution of the geomagnetic field during the Precambrian, more paleointensity data are needed.

@@

\section*{Geologic Settings}
The North American Midcontinent Rift (MCR) is a failed intracontinental rift where protracted magmatic activity lasted from \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}. Midcontinent Rift rocks extensively outcrop in today's Lake Superior region, with the total extent traceable by arcuate magnetic and gravity anomalies that extend to the southwest to Kansas, and to the southeast, to southern Michigan \cite{Hinze2020a}. Previous studies have divided magmatic activity in the rift into four stages based on interpreted changes in relative magmatic volume and the nature of magmatism: early ($\sim$1109–1104 Ma), latent ($\sim$1104–1098 Ma), main ($\sim$1098–1090 Ma) and late ($\sim$1090–1083 Ma) \cite{Vervoort2007a, Heaman2007a, Miller2013a}. In northeastern Minnesota, the Early Gabbro Series and the Felsic Series rocks of the Duluth Complex and reversed-polarity lavas of the lower North Shore Volcanic Group were emplaced during the early stage. The more voluminous Duluth Complex Layered Series and the plagioclase-rich Anorthositic Series, together with an associated $\sim$8 km thick extrusive volcanic sequences of the North Shore Volcanic Group (NSVG), were rapidly emplaced about 10 myr later at \textit{ca.} 1096 Ma during the main stage \cite{Paces1993a, Swanson-Hysell2020a}.

The Beaver Bay Complex, which sits stratigraphically above the Duluth Complex, is another intrusive complex that resulted from main stage magmatism. The exposed area of the Beaver Bay Complex is $\sim$1000 km\textsuperscript{2} where it has been mapped along the northwestern shore of Lake Superior in northeastern Minnesota (Fig. \ref{fig:Geologic_map}). The Beaver Bay Complex is a multi-phase, composite intrusive complex that intrudes parts of the NSVG (Fig. \ref{fig:Geologic_map}; \citeA{Miller1997a, Swanson-Hysell2020a}). Distinct from the deep plutonic intrusions of the Duluth Complex, the majority of the Beaver Bay Complex is formed of hypabyssal intrusions that were emplaced as dikes and sills at shallow depths \cite{Miller1997a}. Most of the Beaver Bay Complex intrusions are dioritic to gabbroic in composition \cite{Miller1997a}. The main lithology of the Beaver River diabase dikes and sills network within the Beaver Bay Complex is an ophitic olivine gabbro (Fig. \ref{fig:Field_photo}), but in wider areas of dikes and the upper parts of thick sills, this rock type can abruptly transition into intergranular olivine oxide gabbro, then into subprismatic (and commonly foliated) ferrogabbro, and finally into granophyric monzodiorite. The more evolved and later emplaced components of the Beaver River diabase network are commonly distinguished as the Silver Bay intrusions in the southern Beaver Bay Complex (Fig. \ref{fig:Geologic_map}). Overall being intermediate in composition, the Silver Bay intrusions lithologies range from ophitic olivine gabbro to ferrogranite \cite{Shank1989a}. Field mapping by \citeA{Miller1994a} found intrusive relationships between the Silver Bay intrusions and the Beaver River diabase. Angular inclusions of the host Beaver River diabase within marginal zones of the Silver Bay intrusions led \citeA{Miller1997a} to interpret that the Silver Bay intrusions intruded after the diabase crystallized.

One distinctive feature of the Beaver River diabase is its inclusions of anorthosite xenoliths. In the southern part of the Beaver Bay Complex, the Beaver River diabase occurs as dikes and sills, typically including anorthosites with various sizes ranging from centimeters to over 150 meters (Figs. \ref{fig:Geologic_map}, \ref{fig:Field_photo}; \citeA{Grout1939a, Morrison1983a}). The diabase in this region intrudes the Palisade rhyolite of the North Shore Volcanic Group (Fig. \ref{fig:Geologic_map}), which has a $^{206}$Pb/$^{238}$U date of 1093.94 $\pm$ 0.28 Ma (2$\sigma$ analytical uncertainty is presented for CA-ID-TIMS dates throughout this work; \citeA{Swanson-Hysell2019a}). The Beaver River diabase is locally intruded by the Silver Bay intrusions (Fig. \ref{fig:Geologic_map}). An aplite unit within the granophyre zone of one of these Silver Bay intrusions has a $^{206}$Pb/$^{238}$U date of 1091.61 $\pm$ 0.14 Ma \cite{Swanson-Hysell2019a}. Another arcuate, sill-like diabase body mapped as the Beaver River diabase outcrops along the eastern part of the complex (Fig. \ref{fig:Geologic_map}; \citeA{Miller1997a}). The diabase composition there is similar to that in the south and it also contains large anorthosite xenoliths with dimensions that exceed 100 meters at Carlton Peak (Fig. \ref{fig:Geologic_map}). The Beaver River diabase in the northern part of the complex, near the Houghtaling Creek area, typically forms narrow, near-vertical dikes instead of sheets in the southern and eastern regions (Fig. \ref{fig:Geologic_map}; \citeA{Miller1994a}). The diabase in this region only locally contains xenoliths of anorthosite.

Hundreds of anorthosite xenoliths have been recognized and mapped within the Beaver River diabase (Fig. \ref{fig:Geologic_map}). Many hill tops in the Beaver Bay Complex, such as Carlton Peak and Britton Peak, are large anorthosite blocks (which lead \citeA{Lawson1893a} to erroneously conclude that they were relict Archean topography). Later work established the anorthosite blocks as xenoliths, which are now extensively documented through geologic mapping of the region (Fig. \ref{fig:Geologic_map}; \citeA{Miller2001a, Miller1988a, Miller1989a, Boerboom2004a, Boerboom2006a, Boerboom2006b, Boerboom2007a}) and outcrop-scale exposures (Fig. \ref{fig:Field_photo}). In the field, the anorthosites typically appear as subrounded to rounded, light-colored, translucent blocks that are in sharp contact with the hosting diabase (Fig. \ref{fig:Field_photo}). They also occur as exposures whose contact with the diabase is covered (Fig. \ref{fig:Field_photo}). \citeA{Grout1939a} suggested that the rounded anorthosites are the result of abrasion during transportation as they were entrained by the diabase (i.e. physical weathering within a magmatic system). While the Beaver River diabase is chilled against the North Shore Volcanic Group lithologies that it intrudes, the diabase is not chilled against the margin of the anorthosite xenoliths \cite{Morrison1983a, Miller1997a}. The lack of chilled contacts is consistent with the anorthosite being at elevated temperatures and cooling at the same time as the diabase magma (Fig. \ref{fig:thermal_history_model}).

The anorthosite xenoliths are dominantly monomineralic plagioclase that has an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}. Interstitial pyroxene and olivine are present in minor concentrations in the xenoliths. Within the Carlton Peak anorthosite xenolith, up to 10 cm oikocrysts of olivine and pyroxene can occur. Nevertheless, the overall olivine content in the anorthosites is low. Interstitial titanomagnetite-ilmenite intergrowths that exceed 100 $\mu$m can be found with microscopy and $<$20 $\mu$m Fe-Ti oxide grains can be detected with scanning electron microscopy (Fig. \ref{fig:Field_photo}). Based on textural differences \citeA{Morrison1983a} divided the anorthosite xenoliths into four groups: one group which typically have well-developed granoblastic texture characterized by equigranular plagioclase crystals; another group which have interlocking, lath-shaped plagioclase crystals; an intermediate group which can have both granoblastic texture and interlocking plagioclase laths; and a brecciated group that have brittle deformation textures superposed on pre-existing textures.

@@

High-precision $^{206}$Pb/$^{238}$U zircon date of 1091.83 $\pm$ 0.21 from one anorthosite in Silver Bay area tightly constraint the age of the diabase intrusion to be 1091.7 $\pm$ 0.2 Ma. previous paleomagnetic directional data from \citeA{Zhang2021a} reveal that vast majority of the anorthosite xenoliths have minimal secondary remanence and single component characteristic remanence magnetizations carried by low-Ti magnetite. Step-wise demagnetization experiments show that the characteristic remanence often unblock sharply between temperature steps 500\textdegree C and 580\textdegree C. Moreover, the site mean paleodirection of the anorthosite xenoliths share a common mean with the directions of their host diabase. Together with thermal conduction modeling, these data are consistent with the interpretation that the anorthosite xenoliths within the Beaver River diabase acquired remanence during cooling with the diabase after their hypabyssal emplacement. A total of  xxx specimens from xx sites were used for paleointensity experiments. The locations of the xx sites are shown in Fig. 1 and the corresponding site mean palaeomagnetic directions from \citeA{Zhang2021a} are shown in Fig. 2(b).

\begin{figure}
\centering
\noindent\includegraphics[width=4.75 in]{Figure/Field_photo.pdf}
\caption{\footnotesize{Field photographs and petrographic images of the Beaver River diabase and the anorthosite xenoliths within it. (A) Centimeter-sized plagioclase megacrysts in the diabase. (B) Rounded anorthosite xenolith with a diameter of $\sim$7 meters fully enclosed within the diabase. (C) Exposure of a giant Carlton Peak anorthosite with a diameter $>$100 m. (D) 27.5 m diameter anorthosite xenolith sampled as paleomagnetic site AX16 and geochronology sample MS99033. (E) Cross polarized (XPL) image of the subophitic texture of diabase at site BD2 (pyroxene partially enclosing plagioclase). (F) XPL image of anorthosite geochronology sample MS99033. Plagioclase crystals exhibit both granoblastic texture and interlocking lath fabrics. (G) Backscattered electron (BSE) image of a large Fe-Ti oxide with titanomagnetite-ilmenite lamellae in Beaver River diabase site BD2. (H) BSE image of micron-sized Fe-Ti oxides exsolved from pyroxene between plagioclase crystals in anorthosite xenolith site AX4. an-plagioclase with $\sim$70\% anorthite; ilm-ilmenite; mag-magnetite; px-pyroxene.}}
\label{fig:Field_photo}
\end{figure}

@@

@@

\bibliography{YZ_ref}
